SubsidyExplorer - Step 0: Data Wrangling

### 2020-03-11 scripts/00_data_wrangling.Rmd
oecd data update

This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

>>>>

# Note: Political entities, dependencies, and naming

The naming of flag and coastal states varies widely across our data sources and we recognize that this can be a politially sensitive matter. For the purposes of merging multiple data sources, we therefore translate all state names from the raw data sources to ISO-3 character codes. Country names are only used for display purposes, and the use of any particular name is not meant to convey any options regarding sovereignty on behalf of the authors.

```{r}
# Load country dependencies helper file
country_dependencies <- read_csv(here::here("data", "lookup-tables", "country_dependencies.csv"))

# EU countries
eu_countries <- country_dependencies$iso3[country_dependencies$is_EU]
eu_country_names <- sort(countrycode(eu_countries, "iso3c", "country.name"))

# EU overseas territories
eu_territories <- country_dependencies$iso3[country_dependencies$is_overseas_territory & country_dependencies$sovereign_iso3 %in% eu_countries]
eu_territory_names <- sort(c(countrycode(eu_territories, "iso3c", "country.name"), "Ascension Island", "Tristan da Cunha"))

# US overseas territories:  American Samoa, Guam, Northern Marianas Islands, Puerto Rico, US Virgin Islands
us_territories <- country_dependencies$iso3[country_dependencies$is_overseas_territory & country_dependencies$sovereign_iso3 == "USA"]
us_territory_names <- sort(countrycode(us_territories, "iso3c", "country.name"))
```

Additionally, it is important to note that the European Union as a political entity is a Member of the World Trade Organization, rather then the individual states that make up the European Union. We therefore need to create a profile for the European Union as a whole.  The following `r length(eu_country_names)` states are considered to be part of the European Union for the purposes of this analysis: `r paste(eu_country_names, collapse = ", ")`.

The Netherlands, France, Denmark, and the United Kingdom also have overseas dependencies that should be considered to be a part of the European Union for statistical purposes. The following overseas dependencies are also considered to be a part of the European Union for the purpose of aggregating data in this analysis: `r paste(eu_territory_names, collapse = ", ")`.

Outside of the EU, the only other dependencies we need to worry about are those that relate to the United States. The following overseas dependencies are considered to be a part of the United States for the purpose of aggregating data in this analysis: `r paste(us_territory_names, collapse = ", ")`.

# Section 1 - Subsidies Data

## Updated estimates and analysis of global fisheries subsidies from Sumaila et al. (2019)

```{r}
### Data Wrangling
# Load and wrangle 2019 data
sumaila_2019_dat <- read_xlsx(sumaila_subsidies_path) %>%
  clean_names(case = "snake") %>%
  mutate(iso3 = case_when(country == "Dominican Rp" ~ "DOM",
                          country == "Micronesia" ~ "FSM",
                          country == "Untd Arab Em" ~ "ARE",
                          TRUE ~ countrycode(country, "country.name", "iso3c"))) %>%
  dplyr::select(iso3, type = class, data_type, value = constant_2018_usd) %>%
  mutate(year = 2018,
         units = "2018 US$",
         source = "Sumaila et al. (2019)") %>%
  arrange(iso3)

# Load lookup table for classification scheme
sumaila_lookup <- read_csv(here::here("data", "lookup-tables", "sumaila_subsidies_classification_lookup_table.csv"))

# Clean up the naming/spelling of the classification scheme as well as data type
sumaila_2019_dat_edit <- sumaila_2019_dat %>%
  left_join(sumaila_lookup, by = "type") %>%
  dplyr::select(iso3, year, category, category_name, type, type_name, value, data_type, units, source)

sumaila_2019_dat_edit$data_type[sumaila_2019_dat_edit$data_type == "not found evidence of subsidy"] <- "Found no evidence of subsidy"

# Deal with the EU
sumaila_2019_dat_eu <- sumaila_2019_dat_edit %>%
  dplyr::filter(iso3 %in% c(eu_countries, eu_territories)) %>%
  mutate(entity = "EU",
         data_type = "Aggregate") %>%
  group_by(entity, year, category, category_name, type, type_name, data_type, units, source) %>%
  summarize(value = sum(value, na.rm = T)) %>%
  rename(iso3 = entity)

# Add the EU back in
sumaila_2019_dat_edit <- sumaila_2019_dat_edit %>%
  bind_rows(sumaila_2019_dat_eu)

# Save
write_csv(sumaila_2019_dat_edit, paste0(data_dir, "sumaila_et_al_2019_subsidies_tidy.csv"))
write_csv(sumaila_2019_dat_edit, paste0(app_data_dir, "sumaila_et_al_2019_subsidies_tidy.csv"))

## Estimates of fisheries support provision by state from the OECD (2008 - 2018)

```{r}
# Load and wrangle OECD data (2009 - 2017)
oecd_fse_dat <- read_csv(oecd_fse_path) %>%
  clean_names(case = "snake") %>%
  dplyr::filter(measure == "USD", # only pull estimates in USD for consistancy
                country != "OECD") %>% # remove totals
  dplyr::select(iso3 = country, code = variable, year, value)

# Load OECD lookup table
oecd_fse_lookup <- read_csv(here::here("data", "lookup-tables", "oecd_fse_classification_lookup_table.csv")) %>%
  distinct(category, category_name, type, type_code, type_name)

# # Create variable for matching 
# oecd_matching <- oecd_lookup %>%
#   mutate(code_for_matching = case_when(!is.na(ssubtype_code) ~ ssubtype_code,
#                                        !is.na(subtype_code) ~ subtype_code,
#                                        !is.na(type_code) ~ type_code))

# Match based on type - that should be plenty for the figure on SubsidyExplorer
oecd_fse_dat_edit <- oecd_fse_dat %>%
  inner_join(oecd_fse_lookup, by = c("code" = "type_code")) %>%
  mutate(data_type = "Reported",
         units = paste0(year, " US$"),
         source = "OECD FSE database") %>%
  dplyr::select(iso3, year, category, category_name, type, type_name, data_type, units, source, value)

# Deal with the EU
oecd_fse_dat_eu <- oecd_fse_dat_edit %>%
  dplyr::filter(iso3 %in% c(eu_countries, eu_territories)) %>%
  mutate(entity = "EU",
         data_type = "Aggregate") %>%
  group_by(entity, year, category, category_name, type, type_name, data_type, units, source) %>%
  summarize(value = sum(value, na.rm = T)) %>%
  rename(iso3 = entity) %>%
  arrange(iso3, year, type)

# Add EU back in 
oecd_fse_dat_edit <- oecd_fse_dat_edit %>%
  bind_rows(oecd_fse_dat_eu)

# Save
write_csv(oecd_fse_dat_edit, paste0(data_dir, "oecd_2019_fse_tidy.csv"))
write_csv(oecd_fse_dat_edit, paste0(app_data_dir, "oecd_2019_fse_tidy.csv"))
